Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-18 paper.tex
Update sweeps perf code.
Some restructuring

In this section we describe the main features of \msprime\ 1.0. Methodologically,
there exists a logical separation between simulating coalescent genealogies and
adding neutral mutations onto these genealogies. Simulating genetic ancestry
requires methods that model the underlying reproductive process (e.g. Moran,
Wright-Fisher, etc.) as well as accounting for recombination of various sorts
(including gene conversion), demography, and selection. These methods are
discussed in the first N subsections below. One of the strengths of the
approach taken in msprime is that neutral mutations can be retrospectively
added to simulated genealogies. Methods to generate mutations, and thereby
generate genetic variation, are therefore discussed in a separate and subsequent
subsection. Finally, we describe the \msprime\ API, which is designed to enforce
the distinction between simulating genealogies and mutations; discuss the
interoperability provided by the use of \tskit\, the tree sequence toolkit;
and highlight the open source development process which allows for \msprime\
to be extended by addition of new features while ensuring robustness and
maintainability of the core codebase.

@@

\begin{figure}
\Large{A FIGURE SHOWING DISCOAL AND MSPRIME}
% \includegraphics[width=\textwidth]{figures/recombination}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}

@@

\subsection*{Data interchange and interoperability}
The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important---if often overlooked---issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

\item The core problem is that the field lacks an efficient mechanism for
    interchanging simulated data, and so developers are forced to choose
    between efficiency and duplication of effort.

\item Side note somewhere: with efficient access to branch length
    statistics~\citep{ralph2019efficiently}, maybe we don't need to
    simulate mutations at all.
[GT: I'm not so sure about this -- the desired statistics/summaries
must be calculated on the observed dataset as well as on the many
simulated datasets.
Unless the observed dataset is stored as a tree sequence,
(because it has been through tsinfer, for example),
it's not clear how branch or node stats could be calculated
from it.
]

\item Tskit has facilited unprecedented interoperability between simulators:
    Forward simulations can also use tree
    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
    tree sequences. Msprime can take tree sequences simulated by these forward
    simulators as input and complete the simulation (``recapitation'').
    See \cite{haller2018tree} for more details.

>>>>

\jkcomment{We should trim this down quite a bit - all we want to do is some basic
roadmapping, like, we discuss the underlying tree sequence object,
then mutations (because it's quite separate from the rest)
and then a bunch of new features for ancestry simulation. There's no
point in listing it out section by section.}
In this section we describe the main features of \msprime\ 1.0. Methodologically,
there exists a logical separation between simulating coalescent genealogies and
adding neutral mutations onto these genealogies. Simulating genetic ancestry
requires methods that model the underlying reproductive process (e.g. Moran,
Wright-Fisher, etc.) as well as accounting for recombination of various sorts
(including gene conversion), demography, and selection. These methods are
discussed in the first N subsections below. One of the strengths of the
approach taken in msprime is that neutral mutations can be retrospectively
added to simulated genealogies. Methods to generate mutations, and thereby
generate genetic variation, are therefore discussed in a separate and subsequent
subsection. Finally, we describe the \msprime\ API, which is designed to enforce
the distinction between simulating genealogies and mutations; discuss the
interoperability provided by the use of \tskit\, the tree sequence toolkit;
and highlight the open source development process which allows for \msprime\
to be extended by addition of new features while ensuring robustness and
maintainability of the core codebase.

\subsection*{Tree sequences}
One of the key reasons for \msprime's significiant performance advantage
over other simulators is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture and the supporting \tskit\
library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we briefly summarise
here.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
}
\end{figure}

\jkcomment{Paragraph describing what a tree sequence is, referring to
Fig.~\ref{fig-ts-example}.}

The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

\item The core problem is that the field lacked an efficient mechanism for
    interchanging simulated data, and so developers were forced to choose
    between efficiency and duplication of effort.

\item Lots of examples of papers where people have been forced to
    create a modified fork of some simulator so that they can get
    access to some internal details, not exposed by the simulators
    interface. Tree sequences provide a complete record of the
    ancestral process, so we can do things like compute IBD
    segments and introgressed tracts directly on the output. See
    also ARG section for recording more information, though.

\item Side note somewhere: with efficient access to branch length
    statistics~\citep{ralph2019efficiently}, maybe we don't need to
    simulate mutations at all. (not in the ABC context, but elsewhere)

\item Tskit has facilited unprecedented interoperability between simulators:
    Forward simulations can also use tree
    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
    tree sequences. Msprime can take tree sequences simulated by these forward
    simulators as input and complete the simulation (``recapitation'').
    See \cite{haller2018tree} for more details.
    (Although, this could be a discussion point?)

@@

\begin{figure}
\includegraphics[width=6cm]{figures/sweeps-perf}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}

@@
